---
title: "Reproducible Research: Peer Assessment 1"
author: Piyush Madan
output: 
  html_document:
    keep_md: true
---
## 1. Loading and preprocessing the data 
###1.1 Set working directory to directory having cloned repo of [Assignment](https://github.com/rdpeng/RepData_PeerAssessment1), extract zip  & read the file
```{r}
setwd("~/Documents/projects/RepData_PeerAssessment1")
data <- read.csv("activity.csv")
summary(data)
head(data)
```
###1.2 Use variables in correct format
```{r}
data$date <- as.Date(data$date)
#data$steps <- as.integer(data$steps)
#data[is.na(data$steps),]$steps <- 0
```

## 2. What is mean total number of steps taken per day?
###2.1 Calculate sum of steps per day
```{r}
library(data.table)
DT <- data.table(data)
#get sum of steps per day
DT_daywise <- DT[, sum(steps), by = date]
``` 
Count of Total steps per day (printing top 10 result)
```{r}
head(DT_daywise,10)
```
##2.2 Histogram of steps taken per day.
```{r}
hist( DT_daywise$V1 , main="Histogram: Count of steps per day", xlab="Steps per day", breaks=50)
```

## 2.3 Finding mean steps per day
Based on days when steps were available (not NA)

*Mean*
```{r}
mean(DT_daywise[DT_daywise$V1!="NA"]$V1)
```  
*Median*
```{r}
median(DT_daywise[DT_daywise$V1!="NA"]$V1)
```  

## What is the average daily activity pattern?
Summary of steps taken (per day)
```{r}
summary(DT_daywise$V1)
```

A time series plot with average steps at given time
```{r}
DT_Average_day <- DT[DT$steps != 0, list(AvgSteps = mean(steps, na.rm = TRUE)), by = interval]

plot.ts(DT_Average_day$AvgSteps, main="Average time series with Average steps in 5 min interval", xlab="5-min interval",ylab="count of steps",type = "l")
```

5-minute interval that, on average, contains the maximum number of steps

```{r}
DT_max_at_interval <- DT[DT$steps != 0, list(MaxSteps = max(steps, na.rm = TRUE)), by = interval]

plot.ts(DT_max_at_interval$MaxSteps, main="Time series plot with Max steps in 5 min interval", xlab="5-min interval",ylab="count of steps",type = "l")
```


## Imputing missing values

###Total no. of missing NAs 
```{r}
length(DT[is.na(DT$steps)]$steps)
```

To imput the missing values (`NA`), we replace it with median value of that interval, if that median value is also `NA`, we check median value for that day, else 0
#DT_imputted <- DT[is.na(DT$steps), list(step = mean(steps, na.rm = FALSE))] # , by = interval]


``` {r}
# imputing 
imputeValues <- function(record) {
  #check if steps are NA for that interval, if yes, replace with median of that particular interval for all available days
  if(is.na(record[1])){
     DT_same_interval_withoutNA_steps <- DT[DT$interval==record[3] & !is.na(DT$steps)]$steps    
    imputted_value_median_by_interval <- median(DT_same_interval_withoutNA_steps)
  #if calculated median is NA, then find median of steps available throughout the day  
    if(is.na(imputted_value_median_by_interval)){    
        DT_same_day_withoutNA_steps <- DT[DT$date==record[2] & !is.na(DT$steps)]$steps     
        imputted_value_median_by_day <- median(DT_same_interval_withoutNA_steps)
        
        if(is.na(imputted_value_median_by_day)) {
          return(0)
       } else {
         return(as.integer(imputted_value_median_by_day))           
       }      
    } else {    
        return(as.integer(imputted_value_median_by_interval))
    }    
  } else {  
    # use original steps if available
   return(as.integer(record[1]))
  }   
}


DT$imputtedSteps <- as.integer(apply(DT, 1, imputeValues))
head(DT)
```

After imputting values,  time series plot with average steps at given time
```{r}
DT_Average_day <- DT[DT$steps != 0, list(AvgSteps = mean(imputtedSteps, na.rm = TRUE)), by = interval]

plot.ts(DT_Average_day$AvgSteps, main="Average time series with Average steps in 5 min interval (After imputting)", xlab="5-min interval",ylab="count of steps",type = "l")
```

After imputting values, 5-minute interval that, on average, contains the maximum number of steps

```{r}
DT_max_at_interval <- DT[DT$steps != 0, list(MaxSteps = max(imputtedSteps, na.rm = TRUE)), by = interval]

plot.ts(DT_max_at_interval$MaxSteps, main="Time series plot with Max steps in 5 min interval (After imputting)", xlab="5-min interval",ylab="count of steps",type = "l")
```

## After imputting, histogram of the total number of steps taken each day after missing values were imputed

```{r}
#DT_daywise$V1[is.na(DT_daywise$V1)] <- 0
summary(DT_daywise)
```

Get sum of steps per day on imputted values
```{r}
DT_daywise_imputted <- DT[, sum(imputtedSteps), by = date]
```
Count of Total steps per day (printing top 10 result)
```{r}
head(DT_daywise_imputted,10)
```
Histogram
```{r}
hist( DT_daywise_imputted$V1 , main="Histogram: Count of steps per day(After imputting)", xlab="Steps per day", breaks=50)
```

## Are there differences in activity patterns between weekdays and weekends?

http://stackoverflow.com/questions/1169248/r-function-for-testing-if-a-vector-contains-a-given-element

Setting weekends column equal to TRUE / FALSE based on date
```{r}
#First setting all dates as FALSE and then Saturday & Sunday as TRUE
DT_daywise$weekend <- FALSE
DT_daywise$weekend[weekdays(as.Date(DT_daywise$date)) %in% c("Saturday","Sunday")] <- TRUE
head(DT_daywise)
```

Summary of steps on weekdays
```{r}
summary( DT_daywise$V1[DT_daywise$weekend])

summary( DT_daywise$V1[!DT_daywise$weekend])
```

###Time series for weekday and weekend 
```{r}
#First setting all dates as FALSE and then Saturday & Sunday as TRUE
DT$weekend <- FALSE
DT$weekend[weekdays(as.Date(DT_daywise$date)) %in% c("Saturday","Sunday")] <- TRUE
DT_weekend <- DT[DT$weekend=="TRUE"]

DT_Average_day_weekend <- DT_weekend[DT$steps != 0, list(AvgSteps = mean(steps, na.rm = TRUE)), by = interval]

plot.ts(DT_Average_day_weekend$AvgSteps, main="Average time series with Average steps in 5 min interval(weekend)", xlab="5-min interval",ylab="count of steps",type = "l")

DT_weekday <- DT[DT$weekend=="FALSE"]

DT_Average_day_weekday <- DT_weekday[DT$steps != 0, list(AvgSteps = mean(steps, na.rm = TRUE)), by = interval]

plot.ts(DT_Average_day_weekday$AvgSteps, main="Average time series with Average steps in 5 min interval(weekday)", xlab="5-min interval",ylab="count of steps",type = "l")
```
